Retrofitting Effect Handlers onto OCaml

### 2020-11-05 retro-concurrency.tex
Updates

\usepackage{balance}
%\usepackage{url}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{xspace}
\usepackage{cleveref}
\usepackage{multirow}
\usepackage{microtype}
\usepackage{soul,xcolor}
\usepackage{mathpartir}
\usepackage{amsmath,amsthm}
\usepackage[utf8]{inputenc}
\input{unicode-preamble.tex}
\usepackage{mathtools}
\usepackage{tikz}
\usetikzlibrary{calc,decorations.pathmorphing,shapes}
\usepackage{stmaryrd}

@@

\newcommand{\mycaption}[1]{\vspace{-4mm}\caption{#1}}

@@

When we first enter OCaml from C, a new fiber is allocated for the main OCaml
stack. Since callbacks may be frequent in OCaml programs that use finalisers,
we run the callbacks on the same fiber as the current one. For example, the
layout of the Multicore OCaml stack at |caml_raise_exn| in the |meander|
example from \S\ref{sec:stack} is shown in Figure~\ref{fig:mcstack}. The
functions |caml_call_c| and |caml_call_ocaml| switch the stacks, and hence are
shown in both the system stack and the fiber. The details of the exception
frames have been elided in the figure. Thanks to the fiber representation,
external calls and callbacks remain fast.

@@

In this section, we measure the impact of the addition of effect handlers on
code that does not use effect handlers.

@@

\section*{Supplementary Material}

@@

\subsection{The impact on code not using effects}

There are several changes to the runtime to support fibers and effect handlers
that will be executed by code not using effects: OCaml stacks now have dynamic
checks in the function prologue and the FFI interface tracks stack switching.

TODO:
\begin{itemize}
\item Aim: |R1| not altered existing code performance too much
\item micro-benchmarks covering 1M external calls, 1M callbacks, 1M install handler return, 1M install handler + raise; 4.10.0+stock vs 4.10.0+multicore
\item sandmark macro benchmarks to show sequential performance: 4.10.0+stock, 4.10.0+multicore,
4.10.0+multicore no red-zone/leaf-optimization, 4.10.0+multicore
\end{itemize}

>>>>

\usepackage{balance}
%\usepackage{url}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{xspace}
\usepackage{cleveref}
\usepackage{multirow}
\usepackage{microtype}
\usepackage{soul,xcolor}
\usepackage{mathpartir}
\usepackage{amsmath,amsthm}
\usepackage[utf8]{inputenc}
\input{unicode-preamble.tex}
\usepackage{mathtools}
\usepackage{tikz}
\usetikzlibrary{calc,decorations.pathmorphing,shapes}
\usepackage{stmaryrd}
\usepackage{adjustbox}

\newcommand{\rot}[1]{\adjustbox{angle=90}{\textbf{#1}}}

@@

\newcommand{\mycaption}[1]{\vspace{-4mm}\caption{#1}}
\setlength{\tabcolsep}{3pt}

@@

When we first enter OCaml from C, a new fiber is allocated for the main OCaml
stack. Since callbacks may be frequent in OCaml programs that use finalisers,
we run the callbacks on the same fiber as the current one. For example, the
layout of the Multicore OCaml stack at |caml_raise_exn| in the |meander|
example from \S\ref{sec:stack} is shown in Figure~\ref{fig:mcstack}. The
functions |caml_call_c| and |caml_call_ocaml| switch the stacks, and hence are
shown in both the system stack and the fiber. Since we are reusing the fiber
for the callback, care must be taken to save and restore the |handler_info|
before calling and after returning from |c_to_ocaml| function, respectively.
The details of the exception frames have been elided in the figure. Thanks to
the fiber representation, external calls and callbacks remain fast.

@@

In this section, we measure the impact of the addition of effect handlers on
code that does not use effect handlers. Our macro benchmark suite consists of
54 real OCaml workloads including verification tools (|Coq|, |Cubicle|,
|AltErgo|), parsers (|menhir|,|yojson|), storage engines (|irmin|), utilities
(|cpdf|,|decompress|), bioinformatics (|fasta|, |knucleotide|, |revcomp2|,
|regexredux2|), numerical analysis (|grammatrix|,|LU_decomposition|) and
simulations (|nbody|,|game_of_life|). In addition to Stock (|stock|) and
Multicore OCaml (|MC|), we also ran the benchmarks on Multicore OCaml with no
red zone (|MC+RedZone0|) in the fiber (all OCaml functions will have stack
overflow check) and a red zone size of 32 words (|MC+RedZone32|). Recall that
the default red zone size in Multicore OCaml is 16 words.

The complete results are presented in the supplementary material, which we
summarise here. On average (geometric mean of the normalized values against
|stock| as the baseline), the multicore variants were \textbf{less than 1\%
slower}. The outliers (on either ends) were due to the difference in the
allocator and the GC between stock and Multicore OCaml. The biggest impact was
the increase in the code size for OCaml code due to stack overflow checks.
Compared to |stock|, the code size for OCaml code 19\% (30\%) more for |MC| and
|MC+RedZone32| (|MC+RedZone0|). The result shows that our red zone is effective
at reducing the code size, but further work is required to bring it closer to
|stock|.

\begin{table}
\caption{Micro benchmarks without effects. Each entry shows percentage
	difference for Multicore OCaml over stock OCaml.}
\vspace{-5mm}
{\small
\begin{tabular}{r c c c c c c c c c c}
	& \rot{exnval} & \rot{exnraise} & \rot{extcall} & \rot{callback} & \rot{ack}
	& \rot{evenodd} & \rot{fib} & \rot{motzkin} & \rot{sudan} & \rot{tak} \\ \hline
	\textbf{Time} & +0.0 & -1.9 & +17 & +65  & +5.3
								& -24 & +2.2 & +10 & +0.0 & +4.2 \\
	\textbf{Instr} & +0.0 & +0.0 & +10 & +72 & +16
								 & +0.0 & +24 & +16 & +14 & +17 \\ \hline
\end{tabular}
}
\label{tab:micro}
\end{table}

We also present micro benchmarks results in Table~\ref{tab:micro}. Since micro
benchmarks magnify micro-architectural optimisations, we also report the number
of instructions executed along with time. |exnval| perform 100 million
iterations of installing exception handlers and returning with value.
|exnraise| is similar, but raises an exception in each iteration. |extcall| and
|callback| perform 100 million external calls and callbacks to identity
functions. The other micro benchmarks are highly recursive programs and were
taken from ~\ref{Folklore}.

The results show that exceptions are no more expensive in |MC| compared to
|stock|. In the other programs, |MC| executes more instructions due to stack
overflow checks, except |evenodd| on which the check is eliminated by the
compiler. Despite this, observe that |MC| is 24\% faster than stock on
|evenodd|. At these scales, we observed that padding tightloops with a few
|nop| instructions, which changes the loop alignment, makes the code upto 15\%
faster. Hence, the difference in running times under 15\% is not statistically
significant. External calls and callbacks are more expensive due to additional
bookkeeping. Callbacks are significantly more expensive, but are also less of a
concern compared to external calls which are far more numerous. As seen
previously, these results however do not affect macro benchmark performance.

@@

\section*{Supplementary Material}
\setcounter{figure}{0}

\subsection*{No effect macro benchmarks}

\begin{figure*}
	\begin{minipage}{\linewidth}
	\includegraphics[width=\linewidth]{sandmark-notebook/sandmark_time}
	\subcaption{Normalized time of macro benchmarks. Baseline is Stock OCaml, whose running time in seconds in given in parenthesis.}
	\includegraphics[width=\linewidth]{sandmark-notebook/sandmark_codesize}
	\subcaption{Normalized OCaml code size of macro benchmarks. Baseline is Stock OCaml, whose size in bytes given in parenthesis.}
	\end{minipage}
	\caption{Macro benchmark performance on code that does not use effects.}
	\label{res:macro}
\end{figure*}

Figure~\ref{res:macro} presents the performance of macro benchmarks that do not
use effects. The aim is to quantify the impact of the addition of effect
handlers to the language on code that does not use effect handlers. The |MC| is
Multicore OCaml with default red zone size (16 words). |MC+RedZone0| uses red
zone size of 0; every OCaml function has a stack overflow check. |MC+RedZone32|
uses a red zone size of 32 words.

@@
